---
title: "Historical Analysis of Novels: Success Across the Centuries"
author: "Madison McCann"
date: "2025-12-08"
output: github_document
---

## Introduction

This report explores the historical and temporal trends ingrained within a curated list of the world's top 500 novels. The primary goal is to understand how the metrics of literary success institutional recognition (library holdings), lasting critical acclaim (average rating), and modern mass appeal (total ratings) have changed over time and across different languages and genres.

The historical analysis focuses on answering: **What defines a "great" novel, and does that definition shift with the century?** We will compare long-term recognition (OCLC Holdings) with short-term popularity (Goodreads Ratings) to critically evaluate the factors contributing to a novel's legacy. This systematic analysis will provide a clear roadmap through the data's historical narrative, culminating in a synthesis of findings on literary endurance.

## Data Description and Preparation 

The analysis is based on the **`top-500-novels-metadata.csv`** dataset, a publicly accessible and highly relevant collection of metadata for 500 novels widely considered to be among the greatest. The data is particularly interesting as it combines traditional measures of influence with modern metrics of engagement.

### Table: Key Variables and Roles

| Variable | Description | Role in Analysis |
| :--- | :--- | :--- |
| `pub_year` | The original publication year of the novel. | Primary temporal variable to track historical trends. |
| `publication_century` | Categorical variable (17th Century and Earlier through 21st Century). | Used for historical grouping and visualization. |
| `genre` | The primary genre of the novel. | Used to analyze popularity distribution across literary categories. |
| `orig_lang` | The original language of publication. | Used to track the dominance and shift of literary cultures over time. |
| `oclc_holdings_clean` | The number of library holdings. | Measure of **historical endurance and institutional importance**. |
| `gr_avg_rating` | Goodreads average user rating. | Measure of **lasting critical quality**. |
| `gr_num_ratings_clean` | Goodreads total number of ratings. | Measure of **modern mass appeal and recent visibility**. |

Data cleaning involved converting the `oclc_holdings` and `gr_num_ratings` from character strings with commas into numeric variables (`oclc_holdings_clean` and `gr_num_ratings_clean`) to facilitate quantitative analysis.

## Reproducibility and Data Cleaning 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#load library
library(tidyverse)
library(ggplot2)

#read the data
df <- read.csv("top-500-novels-metadata_2025-01-11.csv")

#create the categorical publication_century variable
df <- df |>
  mutate(
    publication_century = case_when(
      pub_year <= 1700 ~ "17th Century and Earlier",
      pub_year <= 1800 ~ "18th Century",
      pub_year <= 1900 ~ "19th Century",
      pub_year <= 2000 ~ "20th Century",
      TRUE ~ "21st Century"
    )
  )

#set the correct factor order for the century variable
century_order <- c("17th Century and Earlier", "18th Century", "19th Century", "20th Century", "21st Century")
df$publication_century <- factor(df$publication_century, levels = century_order, ordered = TRUE)

#clean the numeric columns (oclc_holdings and gr_num_ratings)
df <- df |>
  mutate(
    oclc_holdings_clean = as.numeric(gsub(",", "", as.character(oclc_holdings))),
    gr_num_ratings_clean = as.numeric(gsub(",", "", gr_num_ratings))
  )

```

## Main Analysis: Systematic Historical Exploration

### QUESTION 1: How has the distribution of top-ranked novel's publication year changed over time?
This initial analysis establishes the temporal focus of the list, revealing a concentration of literary recognition.

```{r}
century_counts <- df |>
  count(publication_century, name = "novel_count")

print("Novel Counts by Publication Century:")
print (century_counts)

```

The data shows a clear skew toward modern works, with the 20th Century accounting for the majority of the top 500 novels (r twentieth_century_count novels). This suggests that the selection criteria favor relatively recent recognition rather than ancient classics.

```{r}

pub_year_plot <- ggplot(century_counts, aes(x = publication_century, y = novel_count, fill = publication_century)) +
  geom_bar(stat = "identity") +
  geom_text(aes(label = novel_count), vjust = -0.5, size = 3) +
  labs(
    title = "Distribution of Top 500 Novels by Publication Century",
    x = "Publication Century",
    y = "Number of Novels",
    fill = "Century"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "none"
  )

print(pub_year_plot)

ggsave("novels_by_century_bar_char.png", plot = pub_year_plot, width = 7, height = 5)

```

### QUESTION 2: Is there a relationship between a novel's genre and its Goodreads Average Rating?
We explore whether certain genres command higher critical ratings than others among the top-tier novels.

```{r}

top_genres <- df |>
  filter(genre != "") |> 
  count(genre, sort = TRUE)|>
  slice_head(n = 5)|>
  pull(genre)

df_top_genres <- df |>
  filter(genre %in% top_genres)

df_top_genres$genre <- factor(df_top_genres$genre, levels = rev(top_genres))

```

```{r}

genre_rating_plot <- ggplot(df_top_genres, aes(x = gr_avg_rating, y = genre, fill = genre)) +
  geom_boxplot() +
  labs(
    title = "Goodreads Average Rating Distribtuion for Top 5 Genres",
    x = "Goodreads Average Rating",
    y = "Genre",
    fill = "genre"
  ) +
  scale_x_continuous(limits = c(3.5, 4.5), breaks = seq(3.5, 4.5, 0.1)) +
  theme_minimal() +
  theme(legend.position = "none")

print(genre_rating_plot)

ggsave("avg_rating_by_top_genre_boxplot.png", plot = genre_rating_plot, width = 7, height = 5)

```

The box plot shows that while all top genres have high ratings, Bildungsroman and History genres tend to have a higher median and tighter interquartile range, suggesting more consistent critical approval compared to broader genres like Fantasy and Romance, which show a wider range of ratings.

### QUESTION 3: Has the original language of the top novels shifted over time?
This analysis uses a stacked bar chart to identify the historical dominance of different literary cultures.

```{r}

languages_of_interest <- c("English", "French", "Russian", "German", "Spanish")

df_lang_grouped <- df |>
  mutate(
    orig_lang_grouped = case_when(
      orig_lang %in% languages_of_interest ~ orig_lang,
      TRUE ~ "Other"
    )
  )

lang_century_counts <- df_lang_grouped |>
  group_by(publication_century, orig_lang_grouped) |>
  count(name = "novel_count") |>
  ungroup() 

```

```{r}

lang_dominance_plot <- ggplot(lang_century_counts, aes(x = publication_century, y = novel_count, fill = orig_lang_grouped)) +
  geom_bar(stat = "identity", position =  "stack") +
  labs(
    title = "Original Language Distribution of Top Novels by Century",
    subtitle = "English-language novels dominate from the 20th Century onwards.",
    x = "Publication Century",
    y = "Number of Novels",
    fill = "Original Language"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(lang_dominance_plot)

ggsave("lang_dominance_by_century_stacked_bar.png", plot = lang_dominance_plot, width = 8, height = 6)

```

The plot clearly demonstrates that non-English languages (French, Russian, German, Spanish) held significant representation in the 19th Century and earlier, but English-language novels achieve near-total dominance in the 20th and 21st Centuries within this list.

### QUESTION 4: Does the average rating or library holdings of a book correlate with its age?
This section begins the critical verification process by comparing two different metrics of enduring success against the publication year.

Goodreads Average Rating vs. Publication Year
```{r}

age_rating_plot <- ggplot(df, aes(x = pub_year, y = gr_avg_rating)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "purple") +
  labs(
    title = "Goodreads Average Rating vs. Publication Year",
    subtitle = "A slight trend suggests older books may have higher average ratings. ",
    x = "Publication Year",
    y = "Goodreads Average Rating"
  ) + 
  theme_minimal()

print(age_rating_plot)
ggsave("age_vs_avg_rating_scatter.png", plot = age_rating_plot, width = 7, height = 5
       )
```

OCLC Holdings vs. Publication Year
```{r}

age_holdings_plot <- ggplot(df, aes (x = pub_year, y = oclc_holdings_clean)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "red") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "OCLC Holdings vs. Publication Year",
    subtitle = "OCLC Holdings are generally higher for older books, suggesting they have 'stood the test of time'.",
    x = "Publication Year",
    y = "OCLC Holdings (Number of Libraries Holding the Book)"
  ) +
  theme_minimal()

print(age_holdings_plot)
ggsave("age_vs_holdings_scatter.png", plot = age_holdings_plot, width = 7, height = 5)

```

Both gr_avg_rating and oclc_holdings show a negative relationship with publication year (meaning a positive relationship with age). This confirms the hypothesis that novels that have endured the "test of time" tend to be both more widely held by institutions and have a slightly higher average user rating.


### QUESTION 5: How has the number of Goodreads ratings for top-ranked novels changed over time, and what is the distribution of the top-rated novels' publication years by genre?
This adds the third and most current popularity metric (gr_num_ratings_clean) to verify the historical findings and links genre to time.

Goodreads Total Number of Ratings vs. Publication Year
```{r}

age_ratings_count_plot <- ggplot(df, aes(x = pub_year, y = gr_num_ratings_clean)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "darkred") +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Goodreads Total Number of Ratings vs. Publication Year",
    subtitle = "Newer books tend to have a higher count of total ratings, indicating greater recent popularity.",
    x = "Publication Year",
    y = "Goodreads Total Number of Ratings"
  ) +
  theme_minimal()
  
  print(age_ratings_count_plot)
  
  ggsave("scatter_publication_goodreads.png", plot = age_ratings_count_plot, width = 7, height = 5)
```
The trend here is the inverse of Visuals 4A and 4B. Newer books (higher pub_year) have a significantly higher count of total ratings. This highlights a key critical finding: Endurance/Quality (Holdings, Avg Rating) is favored by age, but sheer Volume/Visibility (Total Ratings) is favored by recent publications.



Distribution of Top 5 Genres by Publication Century
```{r}

genre_century_counts <- df_top_genres |>
  group_by(publication_century, genre) |>
  count(name = "novel_count") |>
  ungroup()

genre_century_plot <- ggplot(genre_century_counts, aes(x = publication_century, y = novel_count, fill = genre)) +
  geom_bar(stat = "identity", position = "stack") +
  labs(
    title = "Distribution of Top 5 Genres by Publication Century",
    subtitle = "Certain genres show concentrations in specific centuries.",
    x = "Publication Century",
    y = "Number of Novels",
    fill = "Genre"
  ) +
  scale_fill_brewer(palette = "Set1") + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1)
  )

print(genre_century_plot)

ggsave("distrib_top_genres_bar.png", plot = genre_century_plot, width = 7, height = 5)

```

### QUESTION 6: Is there a correlation between a novel's long-term institutional recognition (OCLC Holdings) and its modern-day popularity (Goodreads Total Ratings)?
This is the final verification step that compares the two strongest measures of success that are independent of publication year, using a formal correlation test and visualization.

Correlation Test
```{r}
correlation_output <- df |>
  summarise(
    correlation = cor(oclc_holdings_clean, gr_num_ratings_clean, use = "complete.obs")
  )
print("Correlation between OCLC Holdings and Goodreads Total Ratings:")
print(correlation_output)

```
The correlation coefficient between OCLC Holdings and Goodreads Total Ratings is 0.276 

OCLC Holdings vs. Goodreads Total Ratings Plot
```{r}

holdings_vs_ratings_plot <- ggplot(df, aes(x = oclc_holdings_clean, y = gr_num_ratings_clean)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", col = "blue") +
  scale_x_continuous(labels = scales::comma) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "OCLC Holdings vs. Goodreads Total Ratings",
    subtitle = "A moderate positive correlation suggests that historically recognized books also achieve modern mass appeal.",
    x = "OCLC Holdings (Number of Libraries Holding the Book)",
    y = "Goodreads Total Number of Ratings"
  ) +
  theme_minimal()

print(holdings_vs_ratings_plot)
# Corrected plot object name:
ggsave("scatter_holdings_goodread_ratings.png", plot = holdings_vs_ratings_plot, width = 7, height = 5)
```

## Conclusion
The analysis of the top 500 novels reveals a shifting narrative about literary greatness. Although the full list shows a significant temporal bias, heavily favoring the 20th and 21st Centuries, the most enduring titles, measured by institutional recognition (OCLC Holdings) and consistent critical quality (Average Goodreads Rating), are often older works. This establishes a critical dichotomy: success defined by enduring quality is distinct from modern success by sheer volume and visibility, which heavily favors newer novels (Total Ratings Count).

Historically, this literary landscape has also undergone a profound cultural shift, marked by the growing dominance of English-language novels from the 20th Century onward, overshadowing previously prominent languages like French and Russian. Ultimately, while certain genres like Bildungsroman and History maintain the most consistently high average ratings, the overall genre representation has changed, with the rise of Fantasy novels.


## Ideas for Future Research
Potential avenues for further research include:

-Nationality vs. Language: Investigate whether non-English language authors are still represented, but through English translations, by analyzing the author_nationality variable.

-The Role of Translation: Analyze the oclc_total_editions variable to see if novels originally published in non-English languages that were translated frequently show a stronger correlation between holdings and modern ratings.

-Decade-Specific Trend Analysis: Use a 10-year set instead of 100-year centuries to generate a smoother, more detailed picture of how metrics like average rating have changed year-by-year.
